\Large Taking uncertainty seriously \newline \normalsize A Bayesian approach to word embedding bias estimation

### 2021-08-06 presentationESSLLI21/presentationESSLLI.Rmd
Final presentation

## The plan of the presentation

- Introduction to the topic of bias in word embeddings
- WEAT and MAC methods
- Methodological problems with MAC implementation
- Limitations of pre-averaging in bias detection methods
- Including uncertainty with Bayesian approach

@@

In the learning process word embeddings can learn implicit biases that reflect harmful stereotypical thinking

@@

Words belonging to an intuitively harmful stereotype are cosine-close to each other

@@

[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]

@@

\footnotesize 
```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("../datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
religion$wordClass <- as.factor(religion$wordClass)
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
set.seed(127)
religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
rownames(religionSmall) <- c()
religionSmall  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "Rows the religion dataset.") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

@@

- Gender-direction might be an indicator of bias, but is insufficient. After debiasing other non-gendered words can remain in biased relations  [@Gonen2019lipstick]

- Methods which involve analogies and their evaluations by human users on Mechanical Turk are unreliable [@Nissim2020fair]

- Sensitive to the choice of protected words, capitalization, distance measures and embedding methods [@zhang2020robustness]

<!-- dodac? -->

@@

## Summary

- Introduction to the topic of bias in word embeddings
- WEAT and MAC methods
- Methodological problems with MAC implementation
- Limitations of pre-averaging in bias detection methods
- Including uncertainty with Bayesian approach

## Further work

>>>>

## Presentation plan

- Bias in word embeddings
- WEAT and MAC methods
- Methodological problems
- Limitations of pre-averaging in bias detection methods
- Accounting for uncertainty with Bayesian approach

@@

Word embeddings can learn implicit harmful biases

@@

Stereotypically connected words are cosine-close

@@

\vspace{2mm}
[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]

@@

\footnotesize 
```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("../datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
religion$wordClass <- as.factor(religion$wordClass)
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
set.seed(127)
religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
rownames(religionSmall) <- c()
religionSmall  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "A few rows from the religion dataset") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

@@

- Gender-direction: insufficient indicator of bias [@Gonen2019lipstick]

- Use of analogies: unreliable [@Nissim2020fair]

- High sensitivity to irrelevant factors [@zhang2020robustness]

@@

## Thank you!

### Summary

- Bias in word embeddings
- WEAT and MAC methods
- Methodological problems
- Limitations of pre-averaging in bias detection methods
- Accounting for uncertainty with Bayesian approach

### Further work

### 2021-08-06 presentationESSLLI21/presentationESSLLI.tex
Final presentation

\begin{frame}{The plan of the presentation}
\protect\hypertarget{the-plan-of-the-presentation}{}
\begin{itemize}
\tightlist
\item
  Introduction to the topic of bias in word embeddings
\item
  WEAT and MAC methods
\item
  Methodological problems with MAC implementation
\item
  Limitations of pre-averaging in biad detection methods
\item
  Bayesian approach
\end{itemize}
\end{frame}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-2}{}
\begin{block}{The worry}
\protect\hypertarget{the-worry}{}
In the learning process word embeddings can learn implicit biases that
reflect harmful stereotypical thinking

@@

\begin{block}{Cosine-based bias: basic intuition}
\protect\hypertarget{cosine-based-bias-basic-intuition}{}
Words belonging to an intuitively harmful stereotype are cosine-close to
each other
\end{block}
\end{frame}

@@

@@

\caption{\label{tab:religionTableHeadEarly}Rows the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llrr}
\toprule
protectedWord & wordToCompare & cosineDistance & cosineSimilarity\\
\midrule
jew & greedy & 0.6947042 & 0.3052958\\
rabbi & greedy & 1.0306175 & -0.0306175\\
rabbi & conservative & 0.7175887 & 0.2824113\\
christian & uneducated & 0.5081939 & 0.4918061\\
christianity & cheap & 1.2816164 & -0.2816164\\
muslim & terrorist & 0.2726106 & 0.7273894\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize
\end{block}
\end{frame}

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-9}{}
\begin{block}{Known challenges}
\protect\hypertarget{known-challenges}{}
\begin{itemize}
\item
  Gender-direction might be an indicator of bias, but is insufficient.
  After debiasing other non-gendered words can remain in biased
  relations (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
  Goldberg, 2019})
\item
  Methods which involve analogies and their evaluations by human users
  on Mechanical Turk are unreliable
  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
\item
  Sensitive to the choice of protected words, capitalization, distance
  measures and embedding methods
  (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
  Stevenson, 2020})
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Summary}
\protect\hypertarget{summary}{}
\begin{itemize}
\tightlist
\item
  Introduced to the topic of bias in word embeddings
\item
  Analysed methodological problems with MAC implementation
\item
  Explained of pre-averaging limitations in methods such as WEAT and MAC
\item
  Proposed Bayesian approach
\end{itemize}
\end{frame}

\begin{frame}{Further work}
\protect\hypertarget{further-work}{}
\begin{itemize}
\tightlist
\item
  Including contrasts in Bayesian calculation
\item
  Performance cross-validation in comparison to other methods (regular
  linear regression, KNN, \dots)
\item
  Downstream tasks and connection with intrinsic evaluation
\item
  Testing data from the original Implicit Association Test (IAT)
\item
  Applying uncertainty to WEAT and better word lists
\item
  Looking at other similarity measures
\end{itemize}
\end{frame}

>>>>

\begin{frame}{Presentation plan}
\protect\hypertarget{presentation-plan}{}
\begin{itemize}
\tightlist
\item
  Bias in word embeddings
\item
  WEAT and MAC methods
\item
  Methodological problems
\item
  Limitations of pre-averaging in bias detection methods
\item
  Accounting for uncertainty with Bayesian approach
\end{itemize}
\end{frame}

@@

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-2}{}
\begin{block}{The worry}
\protect\hypertarget{the-worry}{}
Word embeddings can learn implicit harmful biases

@@

\begin{block}{Cosine-based bias: basic intuition}
\protect\hypertarget{cosine-based-bias-basic-intuition}{}
Stereotypically connected words are cosine-close
\end{block}
\end{frame}

@@

\vspace{2mm}

@@

\caption{\label{tab:religionTableHeadEarly}A few rows from the religion dataset}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llrr}
\toprule
protectedWord & wordToCompare & cosineDistance & cosineSimilarity\\
\midrule
jew & greedy & 0.6947042 & 0.3052958\\
rabbi & greedy & 1.0306175 & -0.0306175\\
rabbi & conservative & 0.7175887 & 0.2824113\\
christian & uneducated & 0.5081939 & 0.4918061\\
christianity & cheap & 1.2816164 & -0.2816164\\
muslim & terrorist & 0.2726106 & 0.7273894\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize
\end{block}
\end{frame}

\begin{frame}{Cosine-based measures of bias}
\protect\hypertarget{cosine-based-measures-of-bias-9}{}
\begin{block}{Known challenges}
\protect\hypertarget{known-challenges}{}
\begin{itemize}
\item
  Gender-direction: insufficient indicator of bias
  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \& Goldberg, 2019})
\item
  Use of analogies: unreliable
  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
\item
  High sensitivity to irrelevant factors
  (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
  Stevenson, 2020})
\end{itemize}
\end{block}
\end{frame}

@@

\begin{frame}{Thank you!}
\protect\hypertarget{thank-you}{}
\begin{block}{Summary}
\protect\hypertarget{summary}{}
\begin{itemize}
\tightlist
\item
  Bias in word embeddings
\item
  WEAT and MAC methods
\item
  Methodological problems
\item
  Limitations of pre-averaging in bias detection methods
\item
  Accounting for uncertainty with Bayesian approach
\end{itemize}
\end{block}

\begin{block}{Further work}
\protect\hypertarget{further-work}{}
\begin{itemize}
\tightlist
\item
  Including contrasts in Bayesian calculation
\item
  Performance cross-validation in comparison to other methods (regular
  linear regression, KNN, \dots)
\item
  Downstream tasks and connection with intrinsic evaluation
\item
  Testing data from the original Implicit Association Test (IAT)
\item
  Applying uncertainty to WEAT and better word lists
\item
  Looking at other similarity measures
\end{itemize}
\end{block}
\end{frame}
